### 2021-10-12 chapters/introduction.tex
removed unwanted files from tracking

    For simulation purposes, the movement of matter can be modelled by the Einstein field equations~Eq.~\ref{eq:Intro:GeqT} and perturbations of black holes~\cite[e.g.,][]{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005}.

@@

    In this case Eq.~\ref{eq:Intro:MultipoleExpansion} evaluates to

@@

    The energy loss due to gravitational waves can be calculated from Eq.~\ref{eq:Intro:MultipoleExpansion} as~\cite{Nagar2005}
     \begin{align}
        \frac{dE}{dt} & = \frac{1}{32 \pi}\sum_{l,m} \left(\left|\frac{\mathrm{d} \mathcal{Q}^{+}_{lm}}{dt}\right|^2 + \left|\mathcal{Q}^\times_{lm}\right|^2\right). \label{eq:Intro:EnergyLoss}
    \end{align}
    For binary systems, this results in an incremental reduction in the orbital separation which was famously measured in the Hulse-Taylor binary neutron star, B1913+16~\cite{Hulse1975,Taylor1982,Weisberg2016}. \par

@@

    Assuming a circular orbit of two point masses of mass $M_1$ and $M_2$, the Newtonian quadrupole formula can be used to calculate the evolution of the orbital separation, $a$, due to gravitational wave emission:~\cite[e.g.,][]{Cutler1994}
    \begin{align}
        \frac{da}{dt} & = -\frac{64}{5}\frac{\mu M^2}{a^3}, \label{eq:Intro:OrbitalEvolution}        
    \end{align}
    where  $M \equiv M_1 + M_2$ is the total mass, and $\mu \equiv M_1 M_2 / M$ is the reduced mass.
    The time evolution of gravitational-wave strain frequency can then be calculated as
    \begin{align}
        \frac{df}{dt} & = \frac{96}{5}\pi^{8/3}\mathcal{M}^{5/3}f^{11/3}, \label{eq:Intro:FreqEvolution}        
    \end{align}   
    where $\mathcal{M} \equiv \mu^{3/5} M^{2/5}$ is the chirp mass of the binary system. 
    Using Eqs.~\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}, an expression for the frequency and the gravitational-wave strain magnitude can be calculated~~\cite[e.g.,][]{Cutler1994}: 
      \begin{align}
      f &  \propto \mathcal{M}^{- 5/8}(t_c-t)^{- 3/8},\label{eq:Intro:chirp_f}\\
        % |h(t)| & \propto  \frac{\pi^{2/3}}{D}\left(\frac{576\mu^3 M^2}{5}\right)^{1/4}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht_old} ,\\
        |h(t)| & \propto  \frac{\mathcal{M}^{5/4}}{D}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht} 
        % Calculated by Cutler1994 (2.12) + (2.11) + Chirp mass defn
    \end{align}
    where $t$ is the time, $t_c$ is the time until coalescence, and $D$ is the luminosity distance to the source. 
    The combination of Eq.~\ref{eq:Intro:chirp_f}~and~\ref{eq:Intro:chirp_ht} result in the chirp waveform detected in the inspiral of compact object mergers.
    \par
    The first direct gravitational-wave detection from the inspiral of a binary neutron star merger, GW170817, was observed by a detector network consisting of the LIGO detectors at Hanford, Washington and Livingston, Louisiana in the USA, and the Virgo detector in Cascina, Italy~\cite{AdvancedLIGO2015,AdvancedVirgo2015,GW170817Detection}. 
    All three detectors consist of `L' shaped laser interferometers with free-falling test-masses.
    The gravitational-wave strain is found by measuring the optical path difference between the perpendicular beam tubes. \par

@@

    % Damour2012 in terms of tidal coupling constant, $\kappa_2^T$
    Gravitational-wave strain emitted from the inspiral of binary neutron star mergers, approximated by Eqs.~\ref{eq:Intro:FreqEvolution}\Hyphdash*\ref{eq:Intro:chirp_ht}, can be recast in the frequency domain using the stationary phase approximation~\cite{Cutler1994}
    \begin{align}
        h_{22}(f) & \equiv \frac{\mathcal{A}}{D}\mathcal{M}^{5/6}f^{-7/6} \exp(i\Psi(f)), \label{eq:Intro:StationaryPhaseApproxH22} \\
        \Psi(f) & = 2\pi f t_c - \phi_c - \frac{\pi}{4} +\frac{3}{4}\left(8\pi\mathcal{M}f\right)^{-5/3}, \label{eq:Intro:StationaryPhaseApproxPsi} 
    \end{align}
    where $h_{22}(f)$ is the Fourier transform of the $(l,m)=(2,2)$ gravitational-wave strain, and $\Psi(f)$ is the corresponding phase.
    Here, $f$ is the frequency~(Eq.~\ref{eq:Intro:FreqEvolution}), $\phi_c$ and $t_c$ are the phase and time of coalescence, and $\mathcal{A}$ is a factor that depends on the orientation of the system.\footnote{The orientation factor, $\mathcal{A}$, depends on angles intrinsic and extrinsic to the binary system. Intrinsic angles include the angles defining the relative orientation of the neutron star spins to the orbital plane. Extrinsic angles include the angles defining the relative orientation between the orbital plane and each of the detectors.}
    Equations~\ref{eq:Intro:StationaryPhaseApproxH22}\Hyphdash*\ref{eq:Intro:StationaryPhaseApproxPsi} are valid for $a \gtrsim 6M$ where $a$ is the orbital separation of the binary system and $a\sim 6M$ is an approximation of the innermost stable circular orbit~\cite{Cutler1994}.
    Solving eqs.~\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution} for $a \sim 6M$ gives $f_{max}\approx c^3 ( 6^{3/2}\pi G M )^{-1}\,\mathrm{Hz}$.
    \par
    % The base post-Newtonian expansion parameter is $\sqrt{M/r}$ but it is more convenient to use $\left(\pi M f\right)^{1/3}$ as the post-Newtonian expansion parameter.
    % These two parameters are equal up to and excluding terms of order $(M/r)^{3/2}$.
    The validity for these equations can be extended to times closer to the merger by adding relativistic corrections calculated by post-Newtonian expansion and effective-one-body dynamics~\cite{Cutler1994,Blanchet1995, Buonanno1999,Droz1999,Damour2012,Yagi2013,Blanchet2014}. % to Eq.~\ref{eq:Intro:StationaryPhaseApproxPsi}
    These corrections introduce dependence on the masses, spins, and tidal deformabilities of the individual neutron stars.
    This allows the measurement of these system properties in addition to the chirp mass, given a gravitational-wave inspiral detection.
    However, the accuracy of the individual neutron star properties is significantly less than the accuracy of the chirp mass because the relativistic corrections required to measure these properties are fairly weak. 
    This can be seen by looking at the parameters measured from GW170817.

@@

    % In comparison, the emission of gravitational waves due to ellipticity and f-modes is approximated by:~\cite{Becker2009}
    % \begin{align}
    %     h_{\mathrm{GW}} \approx 6\times10^{-29} \left(\frac{M}{2.7\,\msun}\right)\left(\frac{R}{10\,\km}\right)^{2}\left(\frac{f_2}{3.3\,\kHz} \right)^2\left(\frac{d}{40\,\Mpc} \right)^{-1}\left(\frac{e}{0.2} \right), \label{eq:Intro:GWFModes}        
    % \end{align}
    % which has been scaled to a distance of $40\,\Mpc$, the distance of GW170817.
    Most postmerger numerical-relativity simulations last for only tens of milliseconds, though some longer simulations have been performed between 50-100\,ms in length~\cite{Rezzolla2010,Ciolfi2017,Ciolfi2019,DePietri2020}.
    % Rezzolla2010 long early simulation showing conservation of M_adm and J_gw
    Although Eq.~\ref{eq:Intro:GWAngularMomentumLoss} shows an estimated emission time for gravitational waves of $\sim 150\,$ms, the uncertainty in the equation of state could place $f_2$ as low as $\sim 2.2\,\kHz$ and some simulations show that the ellipticity reduces over time~\cite{Takami2015,Shibata2017a}.
    Given these possibilities, a four-fold decrease in $e$ together with a reduction in $f_2$ to $2.2\,\kHz$ would result in $t_{\mathrm{GW}} \sim 12\,$s.

>>>>

    For simulation purposes, the movement of matter can be modelled by the Einstein field equations~Eq.(\ref{eq:Intro:GeqT}) and perturbations of black holes~\cite[e.g.,][]{Moncrief1974,Abrahams1996,Andrade1999,Nagar2005}.

@@

    In this case Eq.(\ref{eq:Intro:MultipoleExpansion}) evaluates to

@@

    The energy loss due to gravitational waves can be calculated from Eq.(\ref{eq:Intro:MultipoleExpansion}) as~\cite{Nagar2005}
     \begin{align}
        \frac{dE}{dt} & = \frac{1}{32 \pi}\sum_{l,m} \left(\left|\frac{\mathrm{d} \mathcal{Q}^{+}_{lm}}{dt}\right|^2 + \left|\mathcal{Q}^\times_{lm}\right|^2\right). \label{eq:Intro:EnergyLoss}
    \end{align}
    For binary systems, this results in an incremental reduction in the orbital separation which was famously measured in the Hulse-Taylor binary neutron star, B1913+16~\cite{Hulse1975,Taylor1982,Weisberg2016}. \par

@@

    Assuming a circular orbit of two point masses of mass $M_1$ and $M_2$, the Newtonian quadrupole formula can be used to calculate the evolution of the orbital separation, $a$, due to gravitational wave emission:~\cite[e.g.,][]{Cutler1994}
    \begin{align}
        \frac{da}{dt} & = -\frac{64}{5}\frac{\mu M^2}{a^3}, \label{eq:Intro:OrbitalEvolution}        
    \end{align}
    where  $M \equiv M_1 + M_2$ is the total mass, and $\mu \equiv M_1 M_2 / M$ is the reduced mass.
    The time evolution of gravitational-wave strain frequency can then be calculated as
    \begin{align}
        \frac{df}{dt} & = \frac{96}{5}\pi^{8/3}\mathcal{M}^{5/3}f^{11/3}, \label{eq:Intro:FreqEvolution}        
    \end{align}   
    where $\mathcal{M} \equiv \mu^{3/5} M^{2/5}$ is the chirp mass of the binary system. 
    Using Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}), an expression for the frequency and the gravitational-wave strain magnitude can be calculated~~\cite[e.g.,][]{Cutler1994}: 
      \begin{align}
      f &  \propto \mathcal{M}^{- 5/8}(t_c-t)^{- 3/8},\label{eq:Intro:chirp_f}\\
        % |h(t)| & \propto  \frac{\pi^{2/3}}{D}\left(\frac{576\mu^3 M^2}{5}\right)^{1/4}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht_old} ,\\
        |h(t)| & \propto  \frac{\mathcal{M}^{5/4}}{D}(t_c-t)^{-1/4}, \label{eq:Intro:chirp_ht} 
        % Calculated by Cutler1994 (2.12) + (2.11) + Chirp mass defn
    \end{align}
    where $t$ is the time, $t_c$ is the time until coalescence, and $D$ is the luminosity distance to the source. 
    The combination of Eq.(\ref{eq:Intro:chirp_f})~and~Eq.(\ref{eq:Intro:chirp_ht}) result in the chirp waveform detected in the inspiral of compact object mergers.
    \par
    The first direct gravitational-wave detection from the inspiral of a binary neutron star merger, GW170817, was observed by a detector network consisting of the LIGO detectors at Hanford, Washington and Livingston, Louisiana in the USA, and the Virgo detector in Cascina, Italy~\cite{AdvancedLIGO2015,AdvancedVirgo2015,GW170817Detection}. 
    All three detectors consist of `L' shaped laser interferometers with free-falling test-masses.
    The gravitational-wave strain is found by measuring the optical path difference between the perpendicular beam tubes. \par

@@

    % Damour2012 in terms of tidal coupling constant, $\kappa_2^T$
    Gravitational-wave strain emitted from the inspiral of binary neutron star mergers, approximated by Eqs.(\ref{eq:Intro:FreqEvolution}\Hyphdash*\ref{eq:Intro:chirp_ht}), can be recast in the frequency domain using the stationary phase approximation~\cite{Cutler1994}
    \begin{align}
        h_{22}(f) & \equiv \frac{\mathcal{A}}{D}\mathcal{M}^{5/6}f^{-7/6} \exp(i\Psi(f)), \label{eq:Intro:StationaryPhaseApproxH22} \\
        \Psi(f) & = 2\pi f t_c - \phi_c - \frac{\pi}{4} +\frac{3}{4}\left(8\pi\mathcal{M}f\right)^{-5/3}, \label{eq:Intro:StationaryPhaseApproxPsi} 
    \end{align}
    where $h_{22}(f)$ is the Fourier transform of the $(l,m)=(2,2)$ gravitational-wave strain, and $\Psi(f)$ is the corresponding phase.
    Here, $f$ is the frequency~(Eq.(\ref{eq:Intro:FreqEvolution})), $\phi_c$ and $t_c$ are the phase and time of coalescence, and $\mathcal{A}$ is a factor that depends on the orientation of the system.\footnote{The orientation factor, $\mathcal{A}$, depends on angles intrinsic and extrinsic to the binary system. Intrinsic angles include the angles defining the relative orientation of the neutron star spins to the orbital plane. Extrinsic angles include the angles defining the relative orientation between the orbital plane and each of the detectors.}
    Equations~\ref{eq:Intro:StationaryPhaseApproxH22}\Hyphdash*\ref{eq:Intro:StationaryPhaseApproxPsi} are valid for $a \gtrsim 6M$ where $a$ is the orbital separation of the binary system and $a\sim 6M$ is an approximation of the innermost stable circular orbit~\cite{Cutler1994}.
    Solving Eqs.(\ref{eq:Intro:OrbitalEvolution}\Hyphdash*\ref{eq:Intro:FreqEvolution}) for $a \sim 6M$ gives $f_{max}\approx c^3 ( 6^{3/2}\pi G M )^{-1}\,\mathrm{Hz}$.
    \par
    % The base post-Newtonian expansion parameter is $\sqrt{M/r}$ but it is more convenient to use $\left(\pi M f\right)^{1/3}$ as the post-Newtonian expansion parameter.
    % These two parameters are equal up to and excluding terms of order $(M/r)^{3/2}$.
    The validity for these equations can be extended to times closer to the merger by adding relativistic corrections calculated by post-Newtonian expansion and effective-one-body dynamics~\cite{Cutler1994,Blanchet1995, Buonanno1999,Droz1999,Damour2012,Yagi2013,Blanchet2014}. % to Eq.~\ref{eq:Intro:StationaryPhaseApproxPsi}
    These corrections introduce dependence on the masses, spins, and tidal deformabilities of the individual neutron stars.
    This allows the measurement of these system properties in addition to the chirp mass, given a gravitational-wave inspiral detection.
    However, the accuracy of the individual neutron star properties is significantly less than the accuracy of the chirp mass because the relativistic corrections required to measure these properties are fairly weak. 
    This can be seen by looking at the parameters measured from GW170817.

@@

    % In comparison, the emission of gravitational waves due to ellipticity and f-modes is approximated by:~\cite{Becker2009}
    % \begin{align}
    %     h_{\mathrm{GW}} \approx 6\times10^{-29} \left(\frac{M}{2.7\,\msun}\right)\left(\frac{R}{10\,\km}\right)^{2}\left(\frac{f_2}{3.3\,\kHz} \right)^2\left(\frac{d}{40\,\Mpc} \right)^{-1}\left(\frac{e}{0.2} \right), \label{eq:Intro:GWFModes}        
    % \end{align}
    % which has been scaled to a distance of $40\,\Mpc$, the distance of GW170817.
    Most postmerger numerical-relativity simulations last for only tens of milliseconds, though some longer simulations have been performed between 50-100\,ms in length~\cite{Rezzolla2010,Ciolfi2017,Ciolfi2019,DePietri2020}.
    % Rezzolla2010 long early simulation showing conservation of M_adm and J_gw
    Although Eq.(\ref{eq:Intro:GWAngularMomentumLoss}) shows an estimated emission time for gravitational waves of $\sim 150\,$ms, the uncertainty in the equation of state could place $f_2$ as low as $\sim 2.2\,\kHz$ and some simulations show that the ellipticity reduces over time~\cite{Takami2015,Shibata2017a}.
    Given these possibilities, a four-fold decrease in $e$ together with a reduction in $f_2$ to $2.2\,\kHz$ would result in $t_{\mathrm{GW}} \sim 12\,$s.

### 2021-10-12 papers/paper1.tex
removed unwanted files from tracking

\documentclass[../Thesis.tex]{subfiles}

@@

       \noindent The hats indicate the whitened transformations of the neutron star parameters such that $\widehat{\mathbf{x}}\sim~ \mathcal{N}(0,1)=( \mathbf{x}-\mu_x)/\sigma_x$ where $\mu_x$ and $\sigma_x^2$ are the mean and variance of $\mathbf{x}$ respectively. The compactness parameter can be generated from Eq.~\ref{eq:Ccalc1}-\ref{eq:betacalc1} after determining the values for $a_0, a_1, b_0$, and $b_1$. Spectra can then be trivially generated given any mass, quadrupolar tidal coupling constant and frequency shift. The frequency shift can be determined from the value of the quadrupolar tidal coupling constant~\cite{Bernuzzi2015,Takami2015}. \par

@@

    \section{Results}
        Following the training step, we use Eq.~(\ref{eq:GenerateWaveform1}) to generate spectra.  Figure~\ref{fig:GoodPlotsWithWaveformGeneration} shows how well our generated spectra match the original numerical-relativity spectra. The original spectra are shown as dashed black curves, the cross-validation spectra are shown as red curves, and the one-sigma model uncertainty is shown in shaded light red. All spectra are scaled to a distance of 50 Mpc. The Advanced LIGO noise amplitude spectral density is shown as the dotted black curves~\cite{PSD:aLIGO}.  We fit the large-scale structure of the numerical-relativity peaks well with some deviations in the small-scale structure. \par Figure~\ref{fig:FittingFactorsHistogram} shows a histogram of the noise-weighted fitting factor, Eq.~(\ref{eq:FF1}), between our cross-validated model prediction and the corresponding numerical-relativity spectra for all tested waveforms. The resulting histogram has a mean of 0.95 with a standard deviation of 0.03. \par
        To place the above results in context, we calculate the nearest neighbour fitting factor of the numerical-relativity spectra. We measure the fitting factor for the spectrum under test against all other spectra, and report the largest fitting factor. We obtain nearest neighbour fitting factors for the numerical-relativity spectra of 0.93 with a standard deviation of 0.05. The fitting factors generated by our model compare favourably with this result. Additionally, our model is capable of generating spectra given the required input parameters, whereas a nearest neighbour interpolation would not be capable of this.\par

@@

        Results based on our trained model suggest that the model is selective and could potentially be used in parameter estimation of detected events. If posteriors for the mass and tidal coupling constant are able to be determined, then it is a simple step to calculate the posteriors for the compactness, Eq.~(\ref{eq:Ccalc1}), and the radius of the neutron star. This will be confirmed in future work using a Bayesian framework. Parameter estimation of the post-merger spectra could set bounds on the post-merger quadrupolar tidal coupling constant, allowing comparison with the inspiral value. This could determine whether a phase change in the equation of state is present~\cite{Bauswein2018,Most2018b}.\par

>>>>

% !TeX root = ../Thesis.tex
\documentclass[../Thesis.tex]{subfiles}

@@

       \noindent The hats indicate the whitened transformations of the neutron star parameters such that $\widehat{\mathbf{x}}\sim~ \mathcal{N}(0,1)=( \mathbf{x}-\mu_x)/\sigma_x$ where $\mu_x$ and $\sigma_x^2$ are the mean and variance of $\mathbf{x}$ respectively. The compactness parameter can be generated from Eqs.(\ref{eq:Ccalc1}-\ref{eq:betacalc1}) after determining the values for $a_0, a_1, b_0$, and $b_1$. Spectra can then be trivially generated given any mass, quadrupolar tidal coupling constant and frequency shift. The frequency shift can be determined from the value of the quadrupolar tidal coupling constant~\cite{Bernuzzi2015,Takami2015}. \par

@@

    \section{Results}
        Following the training step, we use Eq.(\ref{eq:GenerateWaveform1}) to generate spectra.  Figure~\ref{fig:GoodPlotsWithWaveformGeneration} shows how well our generated spectra match the original numerical-relativity spectra. The original spectra are shown as dashed black curves, the cross-validation spectra are shown as red curves, and the one-sigma model uncertainty is shown in shaded light red. All spectra are scaled to a distance of 50 Mpc. The Advanced LIGO noise amplitude spectral density is shown as the dotted black curves~\cite{PSD:aLIGO}.  We fit the large-scale structure of the numerical-relativity peaks well with some deviations in the small-scale structure. \par Figure~\ref{fig:FittingFactorsHistogram} shows a histogram of the noise-weighted fitting factor, Eq.(\ref{eq:FF1}), between our cross-validated model prediction and the corresponding numerical-relativity spectra for all tested waveforms. The resulting histogram has a mean of 0.95 with a standard deviation of 0.03. \par
        To place the above results in context, we calculate the nearest neighbour fitting factor of the numerical-relativity spectra. We measure the fitting factor for the spectrum under test against all other spectra, and report the largest fitting factor. We obtain nearest neighbour fitting factors for the numerical-relativity spectra of 0.93 with a standard deviation of 0.05. The fitting factors generated by our model compare favourably with this result. Additionally, our model is capable of generating spectra given the required input parameters, whereas a nearest neighbour interpolation would not be capable of this.\par

@@

        Results based on our trained model suggest that the model is selective and could potentially be used in parameter estimation of detected events. If posteriors for the mass and tidal coupling constant are able to be determined, then it is a simple step to calculate the posteriors for the compactness, Eq.(\ref{eq:Ccalc1}), and the radius of the neutron star. This will be confirmed in future work using a Bayesian framework. Parameter estimation of the post-merger spectra could set bounds on the post-merger quadrupolar tidal coupling constant, allowing comparison with the inspiral value. This could determine whether a phase change in the equation of state is present~\cite{Bauswein2018,Most2018b}.\par
